Clustering Beat-Chroma Patterns in a Large Music Database

### 2010-03-23 Paper/Paper.tex
added emacs latex mode header

@@

\section{Previous Work}\label{sec:prevwork}
Our work can be seen as an extension of the \textit{Shingles} idea
described by Casey and Slaney \cite{Casey2006,Casey2007,Casey2008}.
The two main differences are the clustering framework and the use of
beats and bars to select the position of the \textit{Shingles}.

@@

\section{Data}\label{sec:data}
In this section we present our use of the Echo Nest features and the
datasets we use for training and testing our model.

\subsection{EchoNest API}
We use the Echo Nest analyze API \cite{EchoNest} throughout this work.
The API gives us a chroma vector (length $12$) for every music event (called 
``segment'') for any song we upload to their platform. 
We also get an estimate of where the beats and the bars are. Beats span over
multiple segments, and bars span over multiple beats. 
We transform the chroma vectors per segment into per beat vectors using a 
simple average. We can then stack beat vectors for a number of bars 
(usually 1 or 2). 
We get a fixed-size patch by resampling the number of chroma vectors. Typically,
there are $4$ beats per bar (Figure \ref{fig:barsize}). 
Thus an appropriate pattern size for one bar
is $12$x$4$ or $12$x$8$.

@@

\subsection{Cowbell Dataset}
We have $43,300$ songs, giving  
%$\bf{3,720,091}$ 
$\mathbf{3.7}$\textbf{M} 
non zero bars, a bar usually 
contains $4$ beats. WHERE DOES THIS DATA COMES FROM? GIFT FROM EN?

\subsection{USPOP}
We had access to a low quality (32kbps) version of the songs in the uspop 2002 
data set \cite{uspop2002}.
This well-known data show a great diversity of pop songs, but is not particular
otherwise\footnote{We 
also tried to use Tzanetakis dataset \cite{Tzanetakis2002a}, but 
the $30$ seconds segments seemed hard to analyze by the Echo Nest API, most did
not contain any bar.}.
uspop2002 serves as testing set to measure how well a codebook learned on
the Cowbell data set can represent new songs. We get Echo Nest features
for $8651$ songs of that dataset.

\subsection{Artist Set}
This set encompass $1402$ songs $20$ artists, mostly rock and pop of different 
subgenres. Some results were available regarding the artist recognition
task. Few details given to preserve anonymity.

\section{Algorithm}\label{sec:algo}
In this section we present the creation of the harmonic patterns from
the Echo Nest data. Then, we explain the online vector quantization algorithm
used to learn a codebook of those patterns.

@@

We try to normalize the patterns in respect to the key. We roll the matrix
so the row with the more energy is always the lowest one. This can easily be
seen in Figure \ref{fig:code}. We considered the same technique for rolling
along the other axis and be invariant regarding the downbeat. However, it
seems less robust to noise. Note that we roll each pattern independently.
Also, this implies that the pitch envelope is important rather than the actual 
pitches.

@@

\section{Experiments - Fundamentals}\label{sec:experiments}
In this section we assert the basic properties of our encoding scheme.
The first subsection gives the details to reproduce the experiments,
such as the distance measure and learning rate. In the second subsection, we
measure properties such as ``how the number of codes affect the distortion''
and ``how many learning samples should we use''. Though important numbers
to report, the busy reader could skip to subsection \ref{ssec:random}
and the following sections.
Note that we will present two additional experiments in 
Section \ref{sec:exps2}.

@@

We use the codebook to encode our testing set (see Section \ref{sec:data}).
Each pattern is encoded with only one code. We can measure the average
distance between a pattern and its encoding. We can also measure the use
of the different codes, i.e. the size of the cluster around a code.

@@

\begin{figure}[htb]
\begin{center}
\includegraphics[width=.9\columnwidth]{song_encoded}
\end{center}
\caption{\small{Original song and its encoding. Must get a better one for
the paper!!!!!!}}
\label{fig:encodesong}
\end{figure}

@@

\section{Further Experiments}\label{sec:exps2}
We have shown the basic properties of our clustering algorithm in
Section \ref{sec:experiments}. In Section \ref{sec:visu}, we showed
how to visualize the codebook and the clusters. We know move on to
show other applications that can take advantage of our clustering.
In the next two subsections, we talk about song segmentation and
artist recognition. Our algorithm was not devised for any of these tasks,
but some information contained in the centroids can be naturally used.

@@

\subsection{Artist Recognition}
Let us be clear, the clustering method proposed in this work is not
intended to perform well at artist recognition. An obvious fact is that
we lose all timbral information which would be useful. However, we can
imagine that some artist use more often some patterns than others.
Best accuracy reported\footnote{few details given to preserve anonymity} 
is around $54.5\%$. Our result is $23.4\%$. Random
is around $5\%$. The confusion matrix can be seen in Figure \ref{fig:conf_mat}.
Details of the experiment follow.

@@

@@

\begin{figure}[htb]
\begin{center}
\includegraphics[width=.4\columnwidth]{toriamos_suzannevega_pattern}
\end{center}
\caption{\small{Most typical pattern for Tori Amos and Suzanne Vega.}}
\label{fig:amosvega}
\end{figure}
\fi

@@

Moving on to larger scales, we could imagine merging codes together and
splitting some on two if the clusters become too small or too large.We could
also cluster the codebook itself, in a similar fashion as hierarchical
gaussian mixtures \cite{Vasconcelos2001}.

>>>>

% -*- TeX-PDF-mode: t -*-

@@

\subsection{Previous Work}\label{sec:prevwork}
Our work can be seen as an extension of the \textit{Shingles} idea
described by Casey and Slaney \cite{Casey2006,Casey2007,Casey2008}.
The two main differences are the clustering framework and the use of
beats and bars to select the position of the \textit{Shingles}.

@@

\section{Approach}\label{sec:approach}
In this section we present our use of the Echo Nest features and the
datasets we use for training and testing our model.

\subsection{Features}
We use the Echo Nest analyze API \cite{EchoNest} throughout this work.
The API gives us a chroma vector (length $12$) for every music event (called 
``segment'') for any song we upload to their platform. 
We also get an estimate of where the beats and the bars are. Beats span over
multiple segments, and bars span over multiple beats. 
We transform the chroma vectors per segment into per beat vectors using a 
simple average. We can then stack beat vectors for a number of bars 
(usually 1 or 2). 
We get a fixed-size patch by resampling the number of chroma vectors. Typically,
there are $4$ beats per bar (Figure \ref{fig:barsize}). 
Thus an appropriate pattern size for one bar
is $12$x$4$ or $12$x$8$.

@@

%\section{Algorithm}\label{sec:algo}
In this section we present the creation of the harmonic patterns from
the Echo Nest data. Then, we explain the online vector quantization algorithm
used to learn a codebook of those patterns.

@@

We try to normalize the patterns in respect to the key. We roll the matrix
so the row with the more energy is always the lowest one. This can easily be
seen in Figure \ref{fig:code}. We considered the same techniqueh for rolling
along the other axis and be invariant regarding the downbeat. However, it
seems less robust to noise. Note that we roll each pattern independently.
Also, this implies that the pitch envelope is important rather than the actual 
pitches.

@@

\section{Experiments}\label{sec:experiments}
In this section we assert the basic properties of our encoding scheme.
The first subsection gives the details to reproduce the experiments,
such as the distance measure and learning rate. In the second subsection, we
measure properties such as ``how the number of codes affect the distortion''
and ``how many learning samples should we use''. Though important numbers
to report, the busy reader could skip to subsection \ref{ssec:random}
and the following sections.
Note that we will present two additional experiments in 
Section \ref{sec:exps2}.

\subsection{Training: Cowbell Dataset}\label{sec:traindata}
We have $43,300$ songs, giving  
%$\bf{3,720,091}$ 
$\mathbf{3.7}$\textbf{M} 
non silent bars, a bar usually 
contains $4$ beats. WHERE DOES THIS DATA COMES FROM? GIFT FROM EN?
\footnote{http://www.morecowbell.dj/}

\subsection{Testing: USPOP}\label{sec:testdata}
We had access to a low quality (32kbps) version of the songs in the uspop 2002 
data set \cite{uspop2002}.
This well-known data show a great diversity of pop songs, but is not particular
otherwise\footnote{We 
also tried to use Tzanetakis dataset \cite{Tzanetakis2002a}, but 
the $30$ seconds segments seemed hard to analyze by the Echo Nest API, most did
not contain any bar.}.
uspop2002 serves as testing set to measure how well a codebook learned on
the Cowbell data set can represent new songs. We get Echo Nest features
for $8651$ songs of that dataset.

@@

We use the codebook to encode our testing set (see Section \ref{sec:testdata}).
Each pattern is encoded with only one code. We can measure the average
distance between a pattern and its encoding. We can also measure the use
of the different codes, i.e. the size of the cluster around a code.

@@

\begin{figure}[htb]
\begin{center}
\includegraphics[width=.9\columnwidth]{song_encoded}
\end{center}
\caption{\small{Original song and its encoding. Must get a better one for
the paper!!!!!!  What happened to good day sunshine?}}
\label{fig:encodesong}
\end{figure}

Crazy idea: chord recognition via quantization: transcribe codewords,
then just encode song to get chords. would this work?

@@

\section{Applications}\label{sec:exps2}
We have shown the basic properties of our clustering algorithm in
Section \ref{sec:experiments}. In Section \ref{sec:visu}, we showed
how to visualize the codebook and the clusters. We know move on to
show other applications that can take advantage of our clustering.
In the next two subsections, we talk about song segmentation and
artist recognition. Our algorithm was not devised for any of these tasks,
but some information contained in the centroids can be naturally used.

@@

\subsection{Artist Recognition}

Data:
This set encompass $1402$ songs $20$ artists, mostly rock and pop of different 
subgenres. Some results were available regarding the artist recognition
task. Few details given to preserve anonymity.

Let us be clear, the clustering method proposed in this work is not
intended to perform well at artist recognition. An obvious fact is that
we lose all timbral information which would be useful. However, we can
imagine that some artist use more often some patterns than others.
Best accuracy reported\footnote{few details given to preserve anonymity} 
is around $54.5\%$. Our result is $23.4\%$. Random
is around $5\%$. The confusion matrix can be seen in Figure \ref{fig:conf_mat}.
Details of the experiment follow.

@@

Artists like metallica are characterized by ``wideband'' patterns -
featured muddled due to noisy acoustics - lots of transients,
distortion, etc.  (not by cleaned up by EN's feature analysis)

@@

\begin{figure}[htb]
\begin{center}
\includegraphics[width=.4\columnwidth]{toriamos_suzannevega_pattern}
\end{center}
\caption{\small{Most typical pattern for Tori Amos and Suzanne Vega.
    Can we show the top 2 or 4 for each to get a better idea?}}
\label{fig:amosvega}
\end{figure}
\fi

@@

Moving on to larger scales, we could imagine merging codes together and
splitting some on two if the clusters become too small or too large.We could
also cluster the codebook itself, in a similar fashion as hierarchical
gaussian mixtures \cite{Vasconcelos2001}.
%  (similar idea used for indexing \cite{Muja2009})
